# Measures of Central Tendency and Variability

## Overview
In this chapter, I will quickly go over measures of central tendency and variability.  

## Learning Objectives
By the end of the chapter students should be able to: 

* Solve math problems in R
* Able to assign variables in R
* Know the basic variable types in R

## Measures of Central Tendency

### The Arithmetic mean
The **arithmetic mean** is defined as the sum of the scores divided by the number of scores. It is the average value and is sensitive to outliers. The symbol for sample mean is $\bar{X}$  and the equation form is: 
$$
\bar{X} = \frac{\sum{X_1}}{N} = \frac{X_1 + X_2 + X_3 + ... + X_N}{N}
$$
The calculations for the mean  of the population is the same except the symbol for population is $\mu$ (Greek letter mu). 
$$
\mu = \frac{\sum{X_1}}{N} = \frac{X_1 + X_2 + X_3 + ... + X_N}{N}
$$

#### Properties of the mean
The mean has many important properties and characteristics: 

**1. The mean is sensitive to the exact value of all the scores in the distribution.** To calculate the mean, you need to add all the scores. As such, a change in in any value will change the mean (unlike median and mode).

**2. The sum of the deviations about the equals zero.** Mathematically, this property becomes $\sum(X_i - \bar{X}) = 0$. This states that if the mean is subtracted from each score, the sum of the differences will equal zero. 

### The Overall mean

### The Median

### The Mode

## Measures of Variability 

### The Range

### The Standard Deviation

### The variance 

## Example in R
In your textbook, it asks you to use SPSS to compute the mean, sd, variance, and range for the following set of mathematics exam scores. As well, to name it `Mathexam`. Here I will show you how to do this same task in R. I think you will realize that Descriptive statistics in R is very simple!  

Mathexam: 78, 65, 47, 38, 86, 57, 88, 66, 43, 95, 73, 82, 61

let's start by creating a a variable for these numbers and call it `math_exam`. 
```{r}
math_exam <- c(78, 65, 47, 38, 86, 57, 88, 66, 43, 95, 73, 82, 61)
```

In R, 

* The function `mean()` and `median()` is used to compute the mean and median, respectively,
* The function `sd()` and `var()` is used to compute the standard deviation and variance, respectively. 
* the function `min()`, `max()`, and `range()` is used to compute the minimum value, maximum value, and range, respectively. 

```{r}
# What was the mean exam grade?
mean(math_exam)
# What was median grade?
median(math_exam)
# What was the standard deviation?
sd(math_exam)
# What was the varience?
var(math_exam)
# What was the lowest grade?
min(math_exam)
# What was the highest grade?
max(math_exam)
# What was the lowest and the highest grade?
range(math_exam)
```

Alternatively, you can use the `summary()` function to display several statistic summaries. 

```{r}
summary(math_exam)
```

Or you can use `pastecs::stat.desc()` for even more statistic summaries.
```{r}
library(pastecs)
stat.desc(math_exam)
```
